Commutative character sheaves and geometric types for supercuspidal representations

### 2017-05-04 Geometric types.tex
Catching up

\title[Commutative character sheaves and geometric types]{Commutative character sheaves and geometric types for supercuspidal representations}
\author{Clifton Cunningham}
\address{Department of Mathematics and Statistics, University of Calgary, 2500 University Drive Northwest, Calgary, Alberta, Canada, {T2N~1N4}.}
\email{cunning@math.ucalgary.ca}
\author{David Roe}
\address{Department of Mathematics, University of Pittsburgh, 301 Thackeray Hall, Pittsburgh, PA , United States, 15260.}
\email{roed.math@gmail.com}

@@

\usepackage{amssymb}
\usepackage{amsrefs}
% Fonts
\usepackage{mathrsfs}
% Enumitem
\usepackage{enumitem}
% Hyperrefs
\usepackage{hyperref}

@@

As proved by Ju-Lee Kim in \cite{kim:07a}, all irreducible supercuspidal representations of tamely ramified $p$-adic groups can be built from ``data'' introduced by Jiu-Kang Yu in \cite{yu:01a}*{\S 15}.
While the type, in the sense of Bushnell \& Kutzko \cite{bushnell-kutzko:98a}, of a supercuspidal representation built from Yu data can be constructed directly from the datum, it is convenient to consider and intermediate object, introduced in \cite{yu:01a}*{Remark 15.4}, which we call a \emph{Yu type datum}. 
Yu type data are studied in \cite{Yu:models}, which concludes with the following observation.
%
\begin{quotation}
{\it Therefore, up to some linear characters, all the ingredient representations 
%[appearing in Yu type data]
 are on groups of the form $\underline{H}(\mathcal{O})$, where $\underline{H}$ is a smooth group scheme over $\mathcal{O}$, and the representations are inflated from $\underline{H}(\kappa)$. These results suggest that algebraic geometry and group schemes should play an
important role in the representation theory of $p$-adic groups.} %\cite{Yu:models}
\end{quotation}
%
In this paper we follow the suggestion above by showing that Yu type data are geometrizable, in the following sense.
A Yu type datum determines a sequence of representations $\orho_i$ of compact $p$-adic groups $\oK^i$, for $i=0, \ldots, d$, such that $(\oK^d, \rho_d)$ is a type for a supercuspidal representation of a $p$-adic group.
The main result of \cite{Yu:models} explains how to find, for each $i=0, \ldots, d$, a smooth group scheme $\underline{G}^i$ over the ring $\O$ of integers of the $p$-adic field with $\underline{G}^i(\O)=\,^\circ K^i $.
In this paper we show that if the geometric component group of the reductive quotient of the special fiber of the group scheme $\underline{G}^0$ is cyclic, then each representation $\orho_i$ in this sequence can be replaced by a pair $(\underline{G}^i, \cs{F}^i)$, 
where $\cs{F}^i$ is a virtual Weil sheaf on the Greenberg transform $G^i$ of $\underline{G}^i$.  
Writing $\trFrob{\cs{F}^i}$ for the function on $G^i(k) = \underline{G}^i(\OO{}) = \,^\circ K^i$ obtained by evaluating the trace of the action of Frobenius on the virtual complex $\cs{F}^i$, we show in Theorem~\ref{thm:geotypes} that 
\begin{equation}\label{eqn:intro1}
\trFrob{\cs{F}^i} = \trace(\,^\circ\rho_i).
\end{equation}
By this theorem, then, we obtain geometric avatars for each type in a Yu datum:
\[
\begin{tikzcd}
(\oK^i,\orho_i) \arrow[bend left, dashed]{rrr}{\text{geometrization}} &&& \arrow[bend left, dashed]{lll}{\text{trace of Frob}}
(\underline{G}^i, \cs{F}^i).
\end{tikzcd}
\]
%Theorem~\ref{thm:geotypes} shows how to construct the virtual Weil sheaf $\cs{F}^i$ from the Yu type datum, so that \eqref{eqn:intro1} holds.
We refer to the pair $(\underline{G}^d, \cs{F}^d)$ as a \emph{geometric type}.

@@

We now summarize the sections of the paper in more detail.
%
In Section \ref{sec:defs}, we recall the category $\CS(G)$ from \cite{cunningham-roe:13a} and note that it still makes sense when $G$ is not commutative.  
%
We focus on the case of commutative $G$ in Section \ref{sec:comcom},
giving the definition of a commutative character sheaf and proving our first main theorem, that
$\TrFrob{G} : \CSiso{G} \to G(k)^*$ induces an isomorphism on $\CCSiso{G}$.
%
Passing to the case that $G$ is non-commutative, we give the definition of and main results about commutative character sheaves in Section \ref{sec:noncom}.  
We note that we should only consider character sheaves that arise via pullback from $G\ab$ in order to eliminate those that have nontrivial restriction to the derived subgroup.  
This observation underlies the definition of commutative character sheaves for non-commutative $G$.  
We state our second main result, Theorem~\ref{thm:geo}, that pullback along the abelianization map defines an equivalence of categories $\CCS(G) \to \CCS(G\ab)$.
In Section \ref{ssec:obmor}, we use Galois cohomology to describe the relationship between $G(k)^*$ and $G\ab(k)^*$.  
We also compute the automorphism groups in $\CCS(G)$.
Then in Section \ref{ssec:proofs}, we give proofs of the results in Section \ref{sec:noncom}, which require a development of equivariant linear character sheaves. 
% 
In Section~\ref{sec:types} we use Theorem~\ref{thm:geo} to geometrize types for supercuspidal representations of $p$-adic groups, in a sense made precise in Theorem~\ref{thm:geotypes}. 
% is given in Section~\ref{ssec:geotypes}. 
 %Specifically, we show that if $C$ is a compact $p$-adic group and $\rho$ is a representation of $C$ and $(C,\rho)$ is a type for a tamely ramified $p$-adic group then there is a  group scheme $G$ over $\Fq$ such that $G(\Fq) = C$ and a virtual Weil sheaf $\mathcal{A}$ on $G$ such that $\trFrob{\cs{A}} = \trace(\rho)$.
As preparation for the proof, we review some facts about the Heisenberg-Weil representation and its geometrization, in Section~\ref{ssec:Jacobi}.
Then, in Section~\ref{ssec:review}, we review Yu's theory of types and his study of smooth integral models.  
These elements are pulled together in Section~\ref{ssec:geotypes}, where the proof Theorem~\ref{thm:geotypes} is given.

@@

\begin{lemma}\label{lemma:morinCSab}
If $(\cs{L},\mu,\beta)\in \CS\ab(G)$ then $\mu  : m^*\cs{L} \to \cs{L}\boxtimes\cs{L}$, $\gamma : m^*\cs{L} \to  \theta^*(m^*\cs{L})$ and $\vartheta: \cs{L}\boxtimes\cs{L} \to \theta^*(\cs{L}\boxtimes \cs{L})$ are morphisms in $\CS\ab(G\times G)$.
\end{lemma}
\begin{proof}
Set $G^2 = G\times G$. Define $m^2 : G^2\times G^2 \to G^2$ by $m^2(g_1,g_2,g_1',g_2') = (g_1g_1',g_2g_2')$. Also define $p^2_i : G^2\times G^2 \to G^2$ by $p^2_i(g_1,g_2,g_1,'g_2') = (g_i,g_i')$.
First we show that $m^*\cs{L}$ is an object in $\CS(G^2)$ by equipping it with an isomorphism $\mu_m^2: (m^2)^* (m^*\cs{L}) \to m^*\cs{L} \boxtimes m^*\cs{L}$ defined by the diagram below.
\[
\begin{tikzcd}
(m^2)^* (m^*\cs{L}) \arrow{d}{(m^2)^*\mu} \arrow{rr}{\mu_m^2} && m^*\cs{L} \boxtimes m^*\cs{L}\\
 (m^2)^*(\cs{L}\boxtimes\cs{L}) \arrow[equal]{r} & (m^2)^*(p_1)^*\cs{L} \otimes (m^2)^*(p_2)^*\cs{L} \arrow{r} & (p^2_1)^*m^*\cs{L} \otimes (p^2_2)^*m^*\cs{L} \arrow[equal]{u}
\end{tikzcd}
\] 
The pair $(m^*\cs{L},\mu_m^2)$ satisfies the conditions appearing in \cite{cunningham-roe:13a}*{Def. 1.1}.
The restriction of $m^*\cs{L}$ to $G^2\der= G\der\times G\der$ is canonically isomorphic to $(\EE)_{G^2\der}$ by
\[
\begin{tikzcd}
(m^*\cs{L})\vert_{G^2\der} \arrow{d}{\mu\vert_{G^2\der}} \arrow{r}{\beta^2} & (\EE)_{G^2\der}\\
(\cs{L}\boxtimes\cs{L})\vert_{G^2\der} \arrow[equal]{r} & (\cs{L}\vert_{G\der})\boxtimes(\cs{L}\vert_{G\der}) \arrow{u}{\beta\boxtimes\beta}.
\end{tikzcd}
\]
This shows that $(m^*\cs{L},\mu_m^2,\beta_m^2) \in \CS\ab(G)$. 
Similar work defines $(\cs{L}\boxtimes\cs{L},\mu_\boxtimes^2,\beta_\boxtimes^2) \in \CS\ab(G)$.
By construction, $\mu: m^*\cs{L} \to \cs{L}\boxtimes\cs{L}$ is a morphism in $\CS\ab(G^2)$.
Similar work shows that $\gamma : m^*\cs{L} \to  \theta^*(m^*\cs{L})$ and $\vartheta: \cs{L}\boxtimes\cs{L} \to \theta^*(\cs{L}\boxtimes \cs{L})$ are also morphisms in $\CS\ab(G^2)$.
\end{proof}

@@

Conversely, suppose $(\cs{L},\mu)\in \CS(G)$ and $(\cs{L},\mu) \iso q^*(\cs{L}\ab,\mu\ab)$ in $\CS(G)$.
Then $i^*(\cs{L},\mu) \iso i^*q^*(\cs{L}\ab,\mu\ab)$ in $\CS(G\der)$. 
Since $q\circ i = 1_{G\ab}$, it follows that $\cs{L}\vert_{G\der} \iso (\EE)_{G\der}$ in $\CS(G\der)$.
\end{proof}

>>>>

\title[Commutative character sheaves and geometric types]{Commutative character sheaves and geometric types for supercuspidal representations}
\author{Clifton Cunningham}
\address{Department of Mathematics and Statistics, University of Calgary, 2500 University Drive Northwest, Calgary, Alberta, Canada, {T2N~1N4}.}
\email{cunningham@pims.math.ca}
\author{David Roe}
\address{Department of Mathematics, University of Pittsburgh, 301 Thackeray Hall, Pittsburgh, PA , United States, 15260.}
\email{roed.math@gmail.com}

@@

\usepackage{amssymb}
\usepackage{amsrefs}
% Fonts
\usepackage{mathrsfs}
% Enumitem
\usepackage{enumitem}
% Hyperrefs
\usepackage{hyperref}
\hypersetup{
  colorlinks   = true, %Colours links instead of ugly boxes
  urlcolor     = blue, %Colour for external hyperlinks
  linkcolor    = blue, %Colour of internal links
  citecolor   = green %Colour of citations
}
\usepackage[margin=3.5cm]{geometry}
%\renewcommand{\baselinestretch}{1.0}

@@

As proved by Ju-Lee Kim in \cite{kim:07a}, all irreducible supercuspidal representations of tamely ramified $p$-adic groups can be built from ``data'' introduced by Jiu-Kang Yu in \cite{yu:01a}*{\S 15}.
While the type, in the sense of Bushnell \& Kutzko \cite{bushnell-kutzko:98a}, of a supercuspidal representation built from Yu data can be constructed directly from the datum, it is convenient to consider an intermediate object, introduced in \cite{yu:01a}*{Remark 15.4}, which we call a \emph{Yu type datum}. 
Yu type data are studied in \cite{Yu:models}, which concludes with the following observation.
%
\begin{quotation}
{\it Therefore, up to some linear characters, all the ingredient representations 
%[appearing in Yu type data]
 are on groups of the form $\underline{H}(\mathcal{O})$, where $\underline{H}$ is a smooth group scheme over $\mathcal{O}$, and the representations are inflated from $\underline{H}(\kappa)$. These results suggest that algebraic geometry and group schemes should play an
important role in the representation theory of $p$-adic groups.} %\cite{Yu:models}
\end{quotation}
%
In this paper we follow the suggestion above by showing that Yu type data are geometrizable, in the following sense.
A Yu type datum determines a sequence of representations $\orho_i$ of compact $p$-adic groups $\oK^i$, for $i=0, \ldots, d$, such that $(\oK^d, \rho_d)$ is a type for a supercuspidal representation of a $p$-adic group.
The main result of \cite{Yu:models} explains how to find, for each $i=0, \ldots, d$, a smooth group scheme $\underline{G}^i$ over the ring $\O$ of integers of the $p$-adic field with $\underline{G}^i(\O)=\,^\circ K^i $.
In this paper we show that if the geometric component group of the reductive quotient of the special fiber of the group scheme $\underline{G}^0$ is cyclic, then each representation $\orho_i$ in this sequence can be replaced by a pair $(\underline{G}^i, \cs{F}^i)$, 
where $\cs{F}^i$ is a virtual Weil sheaf on the Greenberg transform $G^i$ of $\underline{G}^i$.  
Writing $\trFrob{\cs{F}^i}$ for the function on $G^i(k) = \underline{G}^i(\OO{}) = \,^\circ K^i$ obtained by evaluating the trace of the action of Frobenius on the virtual complex $\cs{F}^i$, we show in Theorem~\ref{thm:geotypes} that 
\begin{equation}\label{eqn:intro1}
\trFrob{\cs{F}^i} = \trace(\,^\circ\rho_i).
\end{equation}
By this theorem, then, we obtain geometric avatars for each type in a Yu datum:
\[
\begin{tikzcd}
(\oK^i,\orho_i) \arrow[bend left, dashed]{rrr}{\text{geometrization}} &&& \arrow[bend left, dashed]{lll}{\text{trace of Frob}}
(\underline{G}^i, \cs{F}^i).
\end{tikzcd}
\]
%Theorem~\ref{thm:geotypes} shows how to construct the virtual Weil sheaf $\cs{F}^i$ from the Yu type datum, so that \eqref{eqn:intro1} holds.
We refer to the pair $(\underline{G}^d, \cs{F}^d)$ as a \emph{geometric type}.

@@

We now summarize the sections of the paper in more detail.
%
In Section \ref{sec:defs}, we recall the category $\CS(G)$ from \cite{cunningham-roe:13a} and note that it still makes sense when $G$ is not commutative.  
%
We focus on the case of commutative $G$ in Section \ref{sec:comcom},
giving the definition of a commutative character sheaf and proving our first main theorem, that
$\TrFrob{G} : \CSiso{G} \to G(k)^*$ induces an isomorphism on $\CCSiso{G}$.
%
Passing to the case that $G$ is non-commutative, we give the definition of and main results about commutative character sheaves in Section \ref{sec:noncom}.  
We note that we should only consider character sheaves that arise via pullback from $G\ab$ in order to eliminate those that have nontrivial restriction to the derived subgroup.  
This observation underlies the definition of commutative character sheaves for non-commutative $G$.  
We state our second main result, Theorem~\ref{thm:geo}, that pullback along the abelianization map defines an equivalence of categories $\CCS(G) \to \CCS(G\ab)$.
In Section \ref{ssec:obmor}, we use Galois cohomology to describe the relationship between $G(k)^*$ and $G\ab(k)^*$.  
We also compute the automorphism groups in $\CCS(G)$.
Then in Section \ref{ssec:proofs}, we give proofs of the results in Section \ref{sec:noncom}, which require a development of equivariant linear character sheaves. \todo{Fix this.}
% 
In Section~\ref{sec:types} we use Theorem~\ref{thm:geo} to geometrize types for supercuspidal representations of $p$-adic groups, in a sense made precise in Theorem~\ref{thm:geotypes}. 
% is given in Section~\ref{ssec:geotypes}. 
 %Specifically, we show that if $C$ is a compact $p$-adic group and $\rho$ is a representation of $C$ and $(C,\rho)$ is a type for a tamely ramified $p$-adic group then there is a  group scheme $G$ over $\Fq$ such that $G(\Fq) = C$ and a virtual Weil sheaf $\mathcal{A}$ on $G$ such that $\trFrob{\cs{A}} = \trace(\rho)$.
As preparation for the proof, we review some facts about the Heisenberg-Weil representation and its geometrization, in Section~\ref{ssec:Jacobi}.
Then, in Section~\ref{ssec:review}, we review Yu's theory of types and his study of smooth integral models.  
These elements are pulled together in Section~\ref{ssec:geotypes}, where the proof Theorem~\ref{thm:geotypes} is given.

@@

\begin{lemma}\label{lemma:morinCSab}
If $(\cs{L},\mu,\beta)\in \CS\ab(G)$ then $\mu  : m^*\cs{L} \to \cs{L}\boxtimes\cs{L}$, $\gamma : m^*\cs{L} \to  \theta^*(m^*\cs{L})$ and $\vartheta: \cs{L}\boxtimes\cs{L} \to \theta^*(\cs{L}\boxtimes \cs{L})$ are morphisms in $\CS\ab(G\times G)$.
\end{lemma}
\begin{proof}
Set $G^2 = G\times G$. Define $m^2 : G^2\times G^2 \to G^2$ by $m^2(g_1,g_2,g_1',g_2') = (g_1g_1',g_2g_2')$. Also define $p^2_i : G^2\times G^2 \to G^2$ by $p^2_i(g_1,g_2,g_1,'g_2') = (g_i,g_i')$.
First we show that $m^*\cs{L}$ is an object in $\CS(G^2)$ by equipping it with an isomorphism $\mu_m^2: (m^2)^* (m^*\cs{L}) \to m^*\cs{L} \boxtimes m^*\cs{L}$ defined by the diagram below.
\[
\begin{tikzcd}
(m^2)^* (m^*\cs{L}) \arrow{d}{(m^2)^*\mu} \arrow{rr}{\mu_m^2} && m^*\cs{L} \boxtimes m^*\cs{L}\\
 (m^2)^*(\cs{L}\boxtimes\cs{L}) \arrow[equal]{r} & (m^2)^*(p_1)^*\cs{L} \otimes (m^2)^*(p_2)^*\cs{L} \arrow{r} & (p^2_1)^*m^*\cs{L} \otimes (p^2_2)^*m^*\cs{L} \arrow[equal]{u}
\end{tikzcd}
\] 
The pair $(m^*\cs{L},\mu_m^2)$ satisfies the conditions appearing in \cite{cunningham-roe:13a}*{Def. 1.1}.
The restriction of $m^*\cs{L}$ to $G^2\der= G\der\times G\der$ is canonically isomorphic to $(\EE)_{G^2\der}$ by
\[
\begin{tikzcd}
(m^*\cs{L})\vert_{G^2\der} \arrow{d}{\mu\vert_{G^2\der}} \arrow{r}{\beta^2_m} & (\EE)_{G^2\der}\\
(\cs{L}\boxtimes\cs{L})\vert_{G^2\der} \arrow[equal]{r} & (\cs{L}\vert_{G\der})\boxtimes(\cs{L}\vert_{G\der}) \arrow{u}{\beta\boxtimes\beta}.
\end{tikzcd}
\]
This shows that $(m^*\cs{L},\mu_m^2,\beta_m^2) \in \CS\ab(G)$. 
Similar work defines $(\cs{L}\boxtimes\cs{L},\mu_\boxtimes^2,\beta_\boxtimes^2) \in \CS\ab(G)$.
By construction, $\mu: m^*\cs{L} \to \cs{L}\boxtimes\cs{L}$ is a morphism in $\CS\ab(G^2)$.
Similar work shows that $\gamma : m^*\cs{L} \to  \theta^*(m^*\cs{L})$ and $\vartheta: \cs{L}\boxtimes\cs{L} \to \theta^*(\cs{L}\boxtimes \cs{L})$ are also morphisms in $\CS\ab(G^2)$.
\end{proof}

@@

Conversely, suppose $(\cs{L},\mu)\in \CS(G)$ and $(\cs{L},\mu) \iso q^*(\cs{L}\ab,\mu\ab)$ in $\CS(G)$.
Then 
\[
i^*(\cs{L},\mu) \iso i^*q^*(\cs{L}\ab,\mu\ab)
\]
 in $\CS(G\der)$.
Since $q\circ i = 1_{G\ab}$, it follows that $\cs{L}\vert_{G\der} \iso (\EE)_{G\der}$ in $\CS(G\der)$.
\end{proof}
